---
title: "Microarray Data Rspondin KI and WT / E.coli pks+ WT and mutant + DSS - Reanalysis with LIMMA"
output: html_document
---

```{r, results='hide', messages=FALSE}
rm(list=ls())
library(limma)
library(readxl)
library(pheatmap)
library(xtable)
library(reshape2)
```

# Introduction

This is data from mouse experiments performed by Christine Harnack hybridized as dual channel to Agilent 014568 micro array. 

```{r, results='hide', messages=FALSE, warning=FALSE}
# set up the experimental design
exp_design = as.data.frame(read_xlsx("./ExpDesign_2020-05.xlsx",sheet = 1))
colnames(exp_design) = make.names(colnames(exp_design))
rownames(exp_design)=as.character(exp_design$Scan.Name)
exp_design$dye_swap = rep(c(F,T), times=nrow(exp_design)/2)
exp_design$FileName = paste(exp_design$Scan.Name, ".txt", sep="")

# read data from raw files
setwd("../Data/Raw/")

RG =read.maimages(exp_design$FileName,source="agilent.median", columns=list(G="gProcessedSignal",R="rProcessedSignal"))
setwd("../../Code/")
ed = exp_design

# fix outdated chip annotations
new_anno_file = "../Data/Raw/Agilent_14868_annotations_2017-06-27.Rdata"
load(new_anno_file)
old_anno = RG$genes
take_over_cols = colnames(old_anno)[!colnames(old_anno) %in% c("GeneName","Description","SystematicName")]
tmp = old_anno[,take_over_cols]
tmp$index=1:nrow(tmp)
tmp = merge(tmp, anno_tab_14868, by.x="ProbeName", by.y="ProbeID", all.x=T, sort=F)
new_col_order = c(take_over_cols, colnames(tmp)[!colnames(tmp) %in% take_over_cols])
new_anno = tmp[order(tmp$index),new_col_order]

RG$genes = new_anno
```


## Data overview


### Samples 

```{r, results='asis', warning=FALSE}
sel_cols = c("Chip.Barcode","dye_swap","Cy3","Cy5")
tmp = ed[,sel_cols]
print.xtable(xtable(tmp,display=rep("s",ncol(tmp)+1), align=paste("|",rep("l|", ncol(tmp)+1)), file="" , include.rownames=F), type="html")
```

## Raw data

```{r, expr, fig.width=8, fig.height=8}
boxplot(data.frame(log2(RG$G)),main="Green", las=2)
boxplot(data.frame(log2(RG$R)),main="Red", las=2)

plotDensities(RG)
```

# Within-array normalization

```{r, fig.width=8, fig.height=8}
MA <- normalizeWithinArrays(RG, method="loess")
MA.avg <- avereps(MA, ID=MA$genes$ProbeName)

limma::plotMA(MA,array=1,status=ifelse(MA$genes$ControlType==1,RG$genes$ProbeName,"gene"),cex=1, legend=F)
plotDensities(MA, main="Within-array normalized densities")

op <- par(mar = c(14,4,4,2) + 0.1)
boxplot(data.frame(MA$M),main="M", las=2, cex.names=0.6)
boxplot(data.frame(MA$A),main="A", las=2)
par(op)
```

```{r, cor, fig.width=8, fig.height=8}
exp_matrix = MA.avg$M * matrix(rep(ifelse(ed$dye_swap, -1, 1),each=nrow(MA.avg$M)),nrow(MA.avg$M),ncol(MA.avg$M))
dswp = exp_design[colnames(exp_matrix),]$dye_swap
cy3_label = as.character(exp_design[colnames(exp_matrix),]$Cy3)
cy5_label = as.character(exp_design[colnames(exp_matrix),]$Cy5)
array_label = paste(ifelse(dswp, cy3_label, cy5_label), "_",ifelse(dswp,cy5_label, cy3_label),"_",dswp ,sep="")
colnames(exp_matrix) = array_label
cc = cor(exp_matrix,method="spearman")
pheatmap(cc)
```

```{r}
ed$Short = paste(ed$Cy3,ed$Cy5,ed$dye_swap, sep="_")
ed$Condition = paste(ed$Cy3,  sep="_")
ed_orig = ed
```

## Control probes

The following control probes exist on the arrays used in this experiment:

* Corner associated (used for orientation purposes during scanning)
  * Bright corner 
  * Dark corner 
* Negative controls
  * 3xSLv1 (hairpin probe that does not hybridize well with any possible RNA)
* Positive controls
  * Human GAPDH and PGK1 probes
  * Deletion stringency probes (DCP, probe with varying number of insertions/changes with respect to reference; the number after the "_" denotes the number of differences to the reference which should correlate with lower expression)
  * E1A_r60: spike-in probes with concentrations that should cover the whole dynamic range of the array

There are a few other expression probes that are used by Agilent's feature extraction/QC pipeline. 

```{r, qc_probes, fig.width=12, fig.height=3}
control_probes = which(RG$genes$ControlType!=0)
cp_data = RG$G[control_probes,]
cp_names = RG$genes[control_probes,]
selected_controls = ifelse(substr(cp_names$ProbeName,1,4) %in% c("ERCC", "ETG0","ETG1") | substr(cp_names$ProbeName,1,2) == "RC",F,T)

# control probes
for (i in 1:ncol(cp_data)) {
  boxplot(log2(cp_data[selected_controls,i]) ~ factor(cp_names$ProbeName[selected_controls]),las=2, main=paste("Sample",i), outline=F, xlab="")
}
```


```{r}
ed=ed_orig
save(ed, RG, MA, file="../Data/Processed/RspoKI_and_DSS_Colon_micro_array_preprocessed_data.Rdata")
```

# Software versions

```{r}
sessionInfo()
```

